How does the writing system get informative? Differentiation or conservation?

### 2023-07-20 manuscript/wp4.tex
ms 20/07

Is it possible, then, that heterography may be selected for in the cultural evolution of orthography? And if so, what might be the mechanism through which such selection could operate? \textcite[pp.~325--326]{Berg:2021} sketch two models of how a word might enter a state of heterography, as depicted in Fig.~\ref{models_of_heterography}. The first model, \textit{the differentiation model}, explains heterography through the creation of a new orthographic form. For example, many monosyllabic words that are homophonous with function words have tended to adopt alternate spellings (usually by adding material): \textit{be}--\textit{bee}, \textit{but}--\textit{butt}, \textit{by}--\textit{bye}--\textit{buy}, \textit{for}--\textit{fore}--\textit{four}, \textit{in}--\textit{inn}, \textit{or}--\textit{oar}--\textit{ore}, \textit{so}--\textit{sew}, \textit{to}--\textit{too}--\textit{two}, etc. Other cases of heterography appear to be the result of a very conscious choice to disambiguate meanings; for example, in British English, the word \textit{cheque} was introduced to to distinguish the bank order from other meanings of the word \textit{check}, the word \textit{byte} is a deliberate respelling of \textit{bite} to avoid accidental mutation into the closely related term \textit{bit} \parencite{Buchholz:1977}, and the word \textit{lite} is a respelling of \textit{light} that is frequently used in food products to distinguish light-in-weight from light-in-color. Etymological (and folk-etymological) spellings introduced during the Renaissance also sometimes resulted in heterographic homophones, perhaps contributing to their success: \textit{scene}--\textit{seen}, \textit{scent}--\textit{sent}, \textit{whole}--\textit{hole} \parencite[pp.~58--59]{Scragg:1974}. Heterography is also very common among proper nouns: Surnames that are homophonous with ordinary nouns often add <e> (\textit{Clarke}, \textit{Greene}, \textit{Wilde}) or double the final consonant (\textit{Carr}, \textit{Hogg}, \textit{Mann}), and it is common for brand names to adopt creative spellings, such as \textit{flickr}, \textit{Froot Loops}, or \textit{MVMT} (movement).

>>>>

Is it possible, then, that heterography may be selected for in the cultural evolution of orthography? And if so, what might be the mechanism through which such selection could operate? \textcite[pp.~325--326]{Berg:2021} sketch two models of how a word might enter a state of heterography, as depicted in Fig.~\ref{models_of_heterography}. The first model, \textit{the differentiation model}, explains heterography through the creation of a new orthographic forms. For example, the word \textit{lite} is a respelling of \textit{light} frequently used in food products to distinguish light-in-calorific-weight from other meanings of \textit{light}; in British English, the word \textit{cheque} (perhaps influenced by \textit{exchequer}) differentiates the bank draft from other meanings of the word \textit{check}; and the word \textit{byte} was a deliberate respelling of \textit{bite} to avoid accidental mutation into the closely related term \textit{bit} \parencite{Buchholz:1977}. Etymological (and folk-etymological) spellings introduced during the Renaissance also occasionally resulted in heterographic homophones, perhaps contributing to their success: \textit{scene}--\textit{seen}, \textit{scent}--\textit{sent}, \textit{whole}--\textit{hole} \parencite[pp.~58--59]{Scragg:1974}, and many monosyllabic words that are homophonous with common function words have tended to adopt alternate spellings: \textit{be}--\textit{bee}, \textit{but}--\textit{butt}, \textit{by}--\textit{bye}--\textit{buy}, \textit{for}--\textit{fore}--\textit{four}, \textit{in}--\textit{inn}, \textit{or}--\textit{oar}--\textit{ore}, \textit{so}--\textit{sew}, \textit{to}--\textit{too}--\textit{two}, etc. Heterography is also very common in proper nouns: Surnames that are homophonous with common nouns often add <e> (\textit{Clarke}, \textit{Greene}, \textit{Wilde}) or double the final consonant (\textit{Carr}, \textit{Hogg}, \textit{Mann}), and it is common for brand names to adopt creative spellings, such as \textit{flickr}, \textit{Froot Loops}, or \textit{MVMT}.
